---
title: "Final Project"
author: "Keith Williams"
date: "11/23/2016"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
data_file <- '../data/data_6115.txt'
dat <- read_delim(data_file, delim = "\t")
glimpse(dat)
```
# Explore  
```{r}
# histograms of each variable
dat %>% 
    gather("variable", "value") %>% 
    ggplot(aes(value)) +
    geom_histogram() +
    facet_wrap(~variable, nrow = 1, scales = "free_x")
```

So, the response variable `V1` appears to be normally distributed. `V2` looks somewhat uniform, `V3` and `V4` normal but with different variances, `V5` skewed right, looks maybe poisson, and `V6` looks skweed right between 0 and 1.

```{r}
# scatter plots vs V1
dat %>% 
    gather("predictor", "x", 2:6) %>% 
    ggplot(aes(x, V1)) + 
    geom_point() +
    geom_smooth() +
    facet_wrap(~predictor, nrow = 1, scales = 'free_x')
```

Looks like `V2` and `V6` are cubic, while `V3` is quadratic. `V4` and `V5` require more investigation. Is `V5` logarithmic?  
```{r}
ggplot(dat, aes(log(V5))) + geom_histogram()
```

```{r}
ggplot(dat, aes(log(V5), V1)) + geom_point() + geom_smooth()
```

Looks like a clearer relationship after log transformation.

```{r}
# Is V4 dependendent on one of the other variables?
dat %>% 
    gather("variable", "value", V2, V3, V5, V6) %>% 
    ggplot(aes(value, V4)) +
    geom_point() +
    geom_smooth() +
    facet_wrap(~variable, nrow = 1, scales = 'free_x')
```

Interesting pattern with V3

```{r}
synth <- dat %>% 
    mutate(V4_synth = map_dbl(V2, ~rnorm(1, sd = .x))) %>% 
    ggplot(aes(V2, V4_synth)) +
    geom_point() +
    coord_cartesian(ylim = c(-2, 2))

observed <- ggplot(dat, aes(V2, V4)) + geom_point() + coord_cartesian(ylim = c(-2, 2))

gridExtra::grid.arrange(synth, observed)
```

So `V5` is a quadratic function of `V4`. But it's not just a synthesis from V4 + noise, the error is heteroskedastic, increasing with distance from 0.  

```{r}
dat %>% 
    mutate(V5_synth = map_dbl(V4, ~.x^2 + runif(1, max = abs(.x)))) %>% 
    gather("V", "y", V5, V5_synth) %>%  
    ggplot(aes(V4, y)) +
    geom_point() +
    geom_line(aes(y = V4^2), color = "blue") +
    facet_wrap(~V)
```

This isn't quite right, but its close. `V5` is nonnegative, becaue it is the result of a square + some positive random error.

What other correlations exist?  
```{r}
    cormat <- dat %>% 
        cor(use = "complete.obs")
    cormat[upper.tri(cormat)] <- NA
    
    cormat %>% 
        reshape2::melt(na.rm = TRUE) %>%
        ggplot(aes(Var2, Var1, fill = value)) +
        geom_tile(color = 'gray90') +
        scale_fill_gradient2(low = "#67a9cf", high = "#ef8a62", limit = c(-1, 1), name = "correlation") +
        coord_equal() +
        theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
              panel.background = element_blank()) +
        ggtitle("Correlation Matrix")
```

`V3` and `V4` are uncorrelated with the response variable.  
`V2` and `V6` makes sense, since they are both cubic functions

```{r}
ggplot(dat, aes(V2, V6)) + geom_point() + geom_smooth()
```

```{r}
ggplot(dat, aes(V2, V2^2)) + geom_point()
```


### Hypothesis so far  
- `V2` is a random uniform variable between 0 and 1 simulated from `runif(200)`, `V1` is cubic in `V2`.  
- `V3` is a random normal variable with mean 0, sd close to 1, but maybe a little less. `V1` is quadratic in `V2`  
- `V4` is the heteroskedastic error of `V2`, *i.e.* V2 + rnorm(sd = V2)  
- `V5` is a quadratic function of `V4` + some positive error that is a function of `V4`  
- `V6` is a quadratic function of `V2` with no random noise added.  

## Basic Model  
```{r}
mdl <- lm(V1 ~ poly(V6, 3) + poly(V3, 2), data = dat)
summary(mdl)
broom::augment(mdl) %>% 
    ggplot(aes(.fitted, V1)) +
    geom_point()
```

